Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/Virgo Network

### 2013-12-18 doc/Review2Response.txt

The reviewer brings two large objections to our paper: first, that our choice of source population (SNR = 20) is unrealistic (and correspondingly, that our results do not represent “realistic estimates of parameter estimation”), and secondly, that our choice of setting the noise equal to zero is not sufficiently justified (nor why we would prefer an MCMC over a Fisher matrix analysis for such a case).  We discuss both objections in turn, as well as the minor points the reviewer discusses:

Choice of SNR:

The referee is completely correct: phrasing our results in terms of a typical detection scenario is not supported by our selection of SNR=20.  The goal of our study was to provide a quotable estimate for the community when discussing the parameter estimation capabilities of Advanced LIGO, in effect to update the results of previous Fisher matrix studies while correcting for the limitations of the Fisher Matrix when dealing with a non-trivial posterior.

We have made extensive revisions to the text to correct this point. As such, we have modified the text, particularly in the abstract, introduction, and conclusion to make the intent clear.  See new paragraphs 5 and 7 in the introduction, the new sentence in the abstract, and the modified paragraphs 1 and 2 in the conclusion.  Additionally, we have culled any references to “realistic estimates” from the abstract and introduction.  By describing the results as “averaged lower-bound uncertainties”, we believe the information is still physically interesting, while not misleading to the reader.

@@

1. The intro seems to imply that the full exploration of parameter space is not "matched filtering". But of course it is, since it uses Eq. (4)

Our goal was to contrast our parameter estimation with a template bank detection approach.  We have corrected the text to better reflect this.

2. The priors given at the bottom of p. 2 are not astrophysical, but rather conventional. Astrophysical processes would not result into a uniform distribution for masses, and the distribution of systems out to the horizon distance for LIGO is not homogeneous. The zero-spin assumption also needs justification (it is later called a "simplifying assumption", but is it realistic?)

The reviewer is correct that our priors are not astrophysical in nature.  In practice, the observed distribution of neutron star masses from pulsar measurements isn’t sufficiently strong to noticeably influence our recovered posteriors.  As such, we elected to employ an agnostic prior on our neutron star masses.

The zero-spin assumption is not physical, though in practice the maximum possible spin of a neutron star (assuming a reasonable equation of state) is insufficient to induce a large effect in our parameter estimation,

@@

3. In 2.3, noise inner products combine linearly only if noise is uncorrelated among detectors

We have added this to the text .

4. In the caption to Fig. 1, the bit about "reflecting the highest 10% of samples" seems dangerous. What is it about?

At a wall in the parameter space (e.g. q=1), the KDE (which is only employed when generating the plots in Figure 1) needs to be artificially increased so that the smoothing near q=1 does not artificially drop due to the lack of points where q>1.  This was done purely so the smoothed plots would agree with the binned histograms of the same data.  We have clarified the text describing the figure.

@@

5. On p. 9, the 8.5 network threshold is a conventional number; the reality is much more complicated

We have modified the text to describe this as a simplification

### 2013-12-18 doc/paper.tex
fixed everything

We are interested in the posterior $p(\thpara | s)$ as it encodes
information about the prior state of knowledge about the problem in
addition to the likelihood of the signal.  For instance, since it is
assumed that compact binaries form homogeneously in the local universe
observable by advanced detectors, we can safely assume that the
population of sources will be distributed homogeneously in volume, i.e.\
with a density proportional to the square of the luminosity distance $D^2$.  
Our prior information can come from mathematical limits in the parameter space 
or from our \textit{a priori} knowledge of astrophysical systems.  
We employ priors which are
\begin{itemize}
\item uniform in component masses from $0.8M_{\odot} \leq M_{1,2} \leq
  30M_{\odot}$, with a minimum chirp mass (see Eq.~\eqref{eqRatioCM})
  of 0.6 $M_{\odot}$ 
\item uniform in volume, which implies a luminosity distance prior of
  $p(D)dD \propto D^2 dD$,
\item uniform in coalescence time over the segment of data being
  analyzed,
  \item a delta function at zero spin, and
\item isotropic in all angles
\end{itemize}
The mass priors listed here are the generic priors used in follow-up
analysis, not specifically for BNS systems.  While we could have used
a prior with a lower component mass cutoff for the present study, in
practice the higher prior boundary does not affect our current
analysis.  \carl{Note that the choice of a flat prior in component masses is distinct from \cite{Nissanke2013},
 which employed a prior flat in chirp mass and reduced mass-ratio.}

@@

\begin{figure*}[t!]
\centering \includegraphics[trim=7cm 0cm 0cm 0cm,
  clip=true,scale=0.58]{9dpdf}
%\begin{array}{ccc} 
%\includegraphics[width=2.2in]{1dpdfs/mc.pdf} &\includegraphics[width=2.2in]{1dpdfs/eta.pdf}  & \includegraphics[width=2.2in]{1dpdfs/m1.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/time.pdf} &\includegraphics[width=2.2in]{1dpdfs/dist.pdf}  &\includegraphics[width=2.2in]{1dpdfs/iota.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/ra.pdf} & \includegraphics[width=2.2in]{1dpdfs/dec.pdf} & \includegraphics[width=2.2in]{1dpdfs/m2.pdf}
%\end{array}$
\caption{\label{9dPDF} Marginalized 1D  posterior probability densities
  of a typical $1.4M_{\odot}/1.4M_{\odot}$ BNS
  system.  We have plotted each of the 9 parameters for our
  non-spinning BNS problem as stated in section \ref{parameterSection}.
    Note how, even for parameters with excellent recovery, the peak of
  the 1D Gaussian is displaced from the injected value (in dashed
  red).  This is not due to a systematic bias, but is caused by the
  marginalization of a single dimension from the full 9D posterior
  space.  To better see this effect, compare the 1D PDF for
  $\mathcal{M}_{c}$ and $q$ with the 2D marginalized PDF in
  Figure~\ref{1414masses}.  The plots here represent the Gaussian kernel density
  estimator of each PDF.  The estimate for $q$ was 
  computed by reflecting the highest $10\%$ of samples across the $q=1$ boundary.  
  This was done to avoid a false decrease in estimated probability density near
  the boundary.}
\end{figure*}

@@

Given the previous reliance of parameter estimation studies on the Fisher
Information Matrix, it is informative to compare the credible intervals quoted here 
to previously reported values in the literature.  We recompute the credible intervals 
of the chirp mass and the symmetric mass ratio ($\eta \equiv m_1 m_2 / M_{tot}^2$)
at the 68\% level \carl{NO, FIX ME}(1-$\sigma$) for the $1.4M_{\odot}/1.4M_{\odot}$ system presented 
here.  We then compare the findings of the MCMC to the Fisher matrix predictions using
an identical network configuration (HLV), noise curve, and waveform model.  
We compute the Fisher matrix uncertainties 
using a code previously described in \cite{Inadequacies}.  We find that the MCMC credible
 intervals \carl{can} vary significantly from
the estimates of the Fisher matrix, with the Fisher matrix underestimating the 
fractional uncertainties by as much as a factor of \sout{three} \carl{two}.  See Table~\ref{FIMvsMCMC}.  This 
result agrees with the conventional wisdom surrounding the Fisher Matrix accuracy,
as well as results where the Fisher Matrix is computed to higher order \citep{Vitale2010}
   This disagreement, while 
   unsurprising, emphasizes the importance of employing the full parameter estimation
    machinery of an MCMC when making physically relevant claims.

@@

\begin{table}[h!]
\centering
\caption{68\% Credible intervals versus \sout{1-$\sigma$} \carl{$2\sigma$} Fisher matrix uncertainties for mass parameters in each system in the HLV configuration.}
  \tabcolsep=0.11cm
    {\renewcommand{\arraystretch}{1.3} 
\begin{tabular}{lcccccccc}

@@

We use a frequency-domain waveform accurate up to $3.5$ post-Newtonian
(pN) order in phase and 3 pN order in amplitude of the lowest ($l=m=2$) spatial mode.
  We restrict
ourselves to quasi-circular, non-spinning waveforms as a simplifying
assumption.  The standard form of our waveform model, known as the
\textit{TaylorF2} approximant, is calculated via the stationary-phase
approximation.  In this setup, the gravitational-wave amplitude is
given by
\begin{equation}
\tilde{h}(f) = a(t_f) e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $a(t_f)$ is the amplitude evaluated at a stationary-phase
reference point, which to lowest order takes the form $a(t_f) \propto
f^{-7/6} \chmass^{5/6}\Theta(\text{angle})/D$, where $D$ is the
luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M_{tot}\equiv M_1+M_2$, it is convenient to work with the mass
ratio and chirp mass, defined by
\begin{equation}
  q\equiv M_2/M_1~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5}
  M_{tot}^{-1/5},
  \label{eqRatioCM}
\end{equation}
respectively.  Note that most gravitational-wave literature instead uses
  the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 /
  M^2$.  We elect to use $q$ as it is more physically intuitive, and
  because it avoids an integrable singularity that can appear for
  equal-mass systems when employing a prior on $p(\eta)$.  
     By convention, $M_1 \geq M_2$.  The stationary phase then
becomes an expansion in the Newtonian orbital velocity, $v=(\pi
M_{tot} f)^{1/3}$,
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} +
\frac{3}{128}\left(\frac{M_{tot}}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

>>>>

We are interested in the posterior $p(\thpara | s)$ as it encodes
information about the prior state of knowledge about the problem in
addition to the likelihood of the signal.  For instance, since it is
assumed that compact binaries form homogeneously in the local universe
observable by advanced detectors, we can safely assume that the
population of sources will be distributed homogeneously in volume, i.e.\
with a density proportional to the square of the luminosity distance $D^2$.  
Our prior information can come from mathematical limits in the parameter space 
or from our \textit{a priori} knowledge of astrophysical systems.  
We employ priors which are
\begin{itemize}
\item uniform in component masses from $0.8M_{\odot} \leq M_{1,2} \leq
  30M_{\odot}$, with a minimum chirp mass (see Eq.~\eqref{parameterspace})
  of 0.6 $M_{\odot}$ 
\item uniform in volume, which implies a luminosity distance prior of
  $p(D)dD \propto D^2 dD$,
\item uniform in coalescence time over the segment of data being
  analyzed,
  \item a delta function at zero spin, and
\item isotropic in all angles
\end{itemize}
The mass priors listed here are the generic priors used in follow-up
analysis, not specifically for BNS systems.  While we could have used
a prior with a lower component mass cutoff for the present study, in
practice the higher prior boundary does not affect our current
analysis.  \carl{Note that the choice of a flat prior in component masses is distinct from \cite{Nissanke2013},
 which employed a prior flat in chirp mass and reduced mass-ratio.}

@@

\begin{figure*}[t!]
\centering \includegraphics[trim=7cm 0cm 0cm 0cm,
  clip=true,scale=0.58]{9dpdf}
%\begin{array}{ccc} 
%\includegraphics[width=2.2in]{1dpdfs/mc.pdf} &\includegraphics[width=2.2in]{1dpdfs/eta.pdf}  & \includegraphics[width=2.2in]{1dpdfs/m1.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/time.pdf} &\includegraphics[width=2.2in]{1dpdfs/dist.pdf}  &\includegraphics[width=2.2in]{1dpdfs/iota.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/ra.pdf} & \includegraphics[width=2.2in]{1dpdfs/dec.pdf} & \includegraphics[width=2.2in]{1dpdfs/m2.pdf}
%\end{array}$
\caption{\label{9dPDF} Marginalized 1D  posterior probability densities
  of a typical $1.4M_{\odot}/1.4M_{\odot}$ BNS
  system.  We have plotted each of the 9 parameters for our
  non-spinning BNS problem as stated in section \ref{parameterSection}.
    Note how, even for parameters with excellent recovery, the peak of
  the 1D Gaussian is displaced from the injected value (in dashed
  red).  This is not due to a systematic bias, but is caused by the
  marginalization of a single dimension from the full 9D posterior
  space.  To better see this effect, compare the 1D PDF for
  $\mathcal{M}_{c}$ and $q$ with the 2D marginalized PDF in
  Figure~\ref{1414masses}.  The plots here represent the Gaussian kernel density
  estimator of each PDF.  The \sout{estimate} \carl{plot} for $q$ was 
  computed by reflecting the highest $10\%$ of samples across the $q=1$ boundary, \carl{in order
  for the smoothed plot to agree with the binned histograms.}
  \sout{This was done to avoid a false decrease in estimated probability density near
  the boundary.}}
\end{figure*}

@@

Given the previous reliance of parameter estimation studies on the Fisher
Information Matrix, it is informative to compare the credible intervals quoted here 
to previously reported values in the literature.  We recompute the credible intervals 
of the chirp mass and the symmetric mass ratio ($\eta \equiv m_1 m_2 / M_{tot}^2$)
at the 68\% level ($2\times\sigma$) for the $1.4M_{\odot}/1.4M_{\odot}$ system presented 
here.  We then compare the findings of the MCMC to the Fisher matrix predictions using
an identical network configuration (HLV), noise curve, and waveform model.  
We compute the Fisher matrix uncertainties 
using a code previously described in \cite{Inadequacies}.  We find that the MCMC credible
 intervals \carl{can} vary significantly from
the estimates of the Fisher matrix, with the Fisher matrix underestimating the 
fractional uncertainties by as much as a factor of \sout{three} \carl{two}.  See Table~\ref{FIMvsMCMC}.  This 
result agrees with the conventional wisdom surrounding the Fisher Matrix accuracy,
as well as results where the Fisher Matrix is computed to higher order \citep{Vitale2010}
   This disagreement, while 
   unsurprising, emphasizes the importance of employing the full parameter estimation
    machinery of an MCMC when making physically relevant claims.

@@

\begin{table}[h!]
\centering
\caption{68\% Credible intervals versus \carl{$2\times\sigma$} Fisher matrix uncertainties for mass parameters in each system in the HLV configuration.}
  \tabcolsep=0.11cm
    {\renewcommand{\arraystretch}{1.3} 
\begin{tabular}{lcccccccc}

@@

We use a frequency-domain waveform accurate up to $3.5$ post-Newtonian
(pN) order in phase and 3 pN order in amplitude of the lowest ($l=m=2$) spatial mode.
  We restrict
ourselves to quasi-circular, non-spinning waveforms as a simplifying
assumption.  The standard form of our waveform model, known as the
\textit{TaylorF2} approximant, is calculated via the stationary-phase
approximation.  In this setup, the gravitational-wave amplitude is
given by
\begin{equation}
\tilde{h}(f) = a(t_f) e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $a(t_f)$ is the amplitude evaluated at a stationary-phase
reference point, which to lowest order takes the form $a(t_f) \propto
f^{-7/6} \chmass^{5/6}\Theta(\text{angle})/D$, where $D$ is the
luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M_{tot}\equiv M_1+M_2$, it is convenient to work with the mass
ratio and chirp mass, defined in \ref{parameterspace}.
  Note that most gravitational-wave literature instead uses
  the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 /
  M^2$.  We elect to use $q$ as it is more physically intuitive, and
  because it avoids an integrable singularity that can appear for
  equal-mass systems when employing a prior on $p(\eta)$.  
     By convention, $M_1 \geq M_2$.  The stationary phase then
becomes an expansion in the Newtonian orbital velocity, $v=(\pi
M_{tot} f)^{1/3}$,
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} +
\frac{3}{128}\left(\frac{M_{tot}}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}
